---
title: "Social Media, Digital News Consumption and Polarization"
subtitle: "Case Study of the 2018 Ontario General Election"
author: 
  - name: Rafael Campos-Gottardo^[McGill University; rafael.campos-gottardo@mail.mcgill.ca]
    affiliations:
      - name: McGill University
  - name: Simon Kiss^[Wilfrid Laurier University; skiss@wlu.ca]
    affiliations:
      - name: Wilfrid Laurier University 
format: 
  beamer:
    navigation: horizontal
    beameroption: "hide notes"
editor: visual
execute: 
  echo: false
  results: hide
  cache: false
# It is necessary to double backslash the underscore
bibliography: Polarization_Canada.bib
link-citations: true
---

## Headline Findings

- Evidence consistent with a polarizing role for social media usage and digital news consumption in self-reported, observational, cross-sectional data from Ontario 2018
- Primarily occurs amongst voters low in interest in politics
- Contra Dubois and Blank [-@duboisEchoChamberOverstated2018] political interest is positively correlated with affective and policy polarization
@abramowitzPolarizationMyth2008; @fiorinaCultureWarMyth2005; @fiorinaCultureWarMyth2011

\note{this is a test}

## Definition

- **Affective polarization** - Increasing dislike of out-partisans
- **Policy polarization** - Increasing disagreement over  policy issues


## State of the Literature

- Seems like affective polarization in the United States is real @iyengarOriginsConsequencesAffective2019
- Less consensus about policy polarization in the US [@abramowitzPolarizationMyth2008; @fiorinaCultureWarMyth2005; @fiorinaCultureWarMyth2011]
- @merkleyPolarizationEhIdeological2022 and @johnstonAffectivePolarizationCanadian2023 find increases in affective polarization in Canada, less evidence of policy polarization in Canada
- Overall, evidence seems to acquit social media (see [@bakshyExposureIdeologicallyDiverse2015; @boxellInternetCausingPolitical2017; @allcottSocialMediaFake2017])
- Digital news consumption might be different @gainousSurfingPoliticalExtremes2023

## Methods

```{r}
#| label: setup
#| echo: false
#| results: hide
#| output: false

library(lubridate)
library(here)
library(knitr)
opts_knit$set(root.dir=rprojroot::find_rstudio_root_file())
source(here("Code/3_polarization.R"))
source(here("Code/3a_bimodality_coefficients.R"))
source(here("Code/3a_overlapping.R"))

```

-   Ontario Provincial Election Study, produced by [LISPOP](http://lispop.ca)
  - Available for public use at <https://doi.org/10.5683/SP3/VFGL0E> <!-- DOI doesn't work-->
- Online consumer sample from SSI, weighted by gender, age and education 
-   Campaign survey fielded between `r paste(format(min(on18$datestamp),"%B %d"),format(max(on18$datestamp),"%B %d, %Y"), sep="-")`
- Rich set of questions on news consumption and social media usage
  - observational, cross-sectional and self-reports but interesting nonetheless.

## Ontario 2018

- End of the Liberal regime
  - Kathleen Wynne and the party were deeply unpopular
- Doug Ford became PC leader at the last minute
- Ripe grounds for strong feelings about out-partisans

## Sample Validation

```{r}
#| label: tbl-vote-choice
#| tbl-cap: Vote intention of all respondents compared with actual vote results
data.frame(Party=c("Liberal Party of Ontario", "Progressive Conservative Party of Ontario", 
"New Democratic Party of Ontario", 
"Green Party of Ontario"), Vote=c(0.196, 0.405, 0.336, 0.046), Sample=rep("Election", 4))->vote_shares

on18 %>% 
  as_factor() %>%
filter(!is.na(partyvote2018)) %>% 
group_by(partyvote2018) %>% 
summarize(n=n()) %>% 
mutate(Vote=(n/sum(n)), Sample=rep("OPES", nrow(.))) %>% 
select(-n, Party=partyvote2018) %>% 
bind_rows(., vote_shares) %>% 
mutate(Vote=scales::percent(Vote, accuracy=1)) %>% 
pivot_wider(., names_from="Sample", values_from="Vote") %>% 
kable()
```

## Key Independent Variables

- **Primary media source**: respondents were asked how they primarily received news about the 2018 Ontario election 
  - Respondents grouped into **legacy** (TV, radio, print newspapers) **online** (news aggregators, blogs, online newspaper) and **social media** (facebook or twitter feeds); otherwise **mixed**.
- **Social media usage**: How often respondents use social media from **never** to **multiple times a day**
  - No reference to news consumption
  
## Measuring Affective polarization

- Measuring affective polarization is more difficult in multi-party democracies [@wagnerAffectivePolarizationMultiparty2021].
- Therefore, we draw on Wagner's [-@wagnerAffectivePolarizationMultiparty2021] weighted affective polarization (WAP) measure.
- **Calculates the average weighted difference between a respondent's average like score and their like score for each party, weighted by party vote share**
  - Could understate polarization, of low Liberal vote share will weight feelings about the Liberals 


## Measuring Policy Polarization

- Index of 11 policy items (see Appendix in the paper)
- Two measures of policy polarization ( see [@lelkesMassPolarizationManifestations2016; @levenduskyRedStatesVs2011]
  - bimodality coefficient to examine bimodality of a distribution of policy preferences [@pfisterGoodThingsPeak2013]
  - the overlap coefficient to examine overlap between two empirical distributions of policy preferences [@pastoreMeasuringDistributionSimilarities2019]  

<!-- <!--:::: {.columns} -->

<!-- ::: {.column width="50%"} -->

<!-- **Bimodality Coefficient** -->

<!-- $$ -->
<!-- BC = \frac{s^2 + 1}{k + 3 *\frac{(n-1)^2}{(n-2)(n-3)}} -->
<!-- $$ -->

<!-- ::: -->

<!-- ::: {.column width="50%"} -->
<!-- **Overlap Coefficent** -->
<!-- $$ -->
<!-- OVL = \int^{+\infty}_{-\infty}|f(x) - g(x)|dx -->
<!-- $$ -->
<!-- ::: -->

<!-- ::::-->

## Affective Polarization and Primary Media Source

```{r}
#| fig-height: 8
WAP_models %>%
  filter(Model_Name %in% c("Model 1", "Model 2", "Model 3", "Model 14")) %>%
  filter(Variable=="Primary_media") %>%
  #Filter out the interaction model
  #filter(Model_Name!="Model 8") %>%
  filter(str_detect(term, "Intercept", negate=T))  %>%
  ggplot(., aes(x=estimate, y=fct_relevel(term, "Social_Media", "Online", "Mixed", after=10)))+facet_wrap(~~fct_relevel(Model_Name, "Model 14",  after=3))+
  geom_pointrange(aes(xmin=estimate-(1.96*std.error), xmax=estimate+(1.96*std.error)))+geom_vline(xintercept=0, linetype=2, col="red")+labs(y="Variable", x="Coefficient") + theme_bw(base_size = 24)
```


## Affective Polarization and Social Media Usage

```{r}
#| fig-height: 8
WAP_models %>%
    filter(Model_Name %in% c("Model 1", "Model 2", "Model 3", "Model 14")) %>%
  filter(Variable=="Social_Use2") %>%
    #filter(Model_Name!="Model 12") %>%
  filter(str_detect(term, "Intercept", negate=T)) %>%
  ggplot(., aes(x=estimate, y=fct_relevel(term, "Never", "Several times a day","About once a day","Several times a week","About once a week", "Less than once a week", after=10)))+facet_wrap(~fct_relevel(Model_Name, "Model 14",  after=3))+
  geom_pointrange(aes(xmin=estimate-(1.96*std.error), xmax=estimate+(1.96*std.error)))+geom_vline(xintercept=0, linetype=2, col="red")+labs(y="Variable", x="Coefficient") + theme_bw(base_size=24)
```
## Notable Trends in Social Media Usage and Affective Polarization

-  Online news consumption is associated with lower levels of affective polarization than legacy media consumption.
- **But** using social media several times a day is associated with higher levels of affective polarization than never using social media.
- Interest is positively related to levels of affective polarization, *contra* 


## Policy Polarization and Primary Media Source - Bimodality Coefficient


:::: {.columns}

::: {.column width="50%"}
```{r}
library(gt)
on18 %>%
  group_by(Primary_media) %>%
  summarize(Bimodality=bimodality_coefficient(policy_polarization)) %>%
   mutate(Bimodality=round(Bimodality,2)) %>%
  gt()
```

:::

::: {.column width="50%"}
```{r}

on18 %>%
  group_by(Interest_cat,Primary_media, ) %>%
  filter(!is.na(Interest_cat)) %>%
  summarize(Bimodality=bimodality_coefficient(policy_polarization)) %>%
rename(Interest=Interest_cat) %>%
  mutate(Bimodality=round(Bimodality,2)) %>%
gt()
```

:::

::::

## Policy Polarization and Social Media Usage - Bimodality Coefficient


```{r}

on18 %>%
  group_by(Social_Use2) %>%
  summarize(Bimodality=bimodality_coefficient(policy_polarization)) %>%
  filter(!is.na(Social_Use2)) %>%
#arrange(., desc(Bimodality)) %>%
  gt() %>%
fmt_number(., columns=2, decimals=2)
```



## Policy Polarization - Overlap Coefficent

:::: {.columns}

::: {.column width="50%"}
```{r}
ggpubr::ggarrange(overlap_legacy, overlap_online, overlap_smedia, overlap_mixed)
```

:::

::: {.column width="50%"}
```{r}
ggpubr::ggarrange(overlap_often, overlap_rarely, ncol = 1)
```

:::

::::

## Discussion & Conclusion


## References {.allowframebreaks}
\footnotesize
