---
title: "Social Media, Digital News Consumption and Polarization"
subtitle: "Case Study of the 2018 Ontario General Election"
author: 
  - name: Rafael Campos-Gottardo^[McGill University; rafael.campos-gottardo@mail.mcgill.ca]
    affiliations:
      - name: McGill University
  - name: Simon Kiss^[Wilfrid Laurier University; skiss@wlu.ca]
    affiliations:
      - name: Wilfrid Laurier University 
format: 
  beamer:
    navigation: horizontal
    beameroption: "hide notes"
editor: visual
execute: 
  echo: false
  results: hide
  cache: false
pdf-engine: lualatex
# It is necessary to double backslash the underscore
bibliography: Polarization_Canada.bib
link-citations: true
include-in-header:
      - text: |
          \usepackage{emoji}
---



## Methods

```{r}
#| label: setup
#| echo: false
#| results: hide
#| output: false

library(lubridate)
library(here)
library(knitr)
opts_knit$set(root.dir=rprojroot::find_rstudio_root_file())
source(here("Code/3_polarization.R"))
source(here("Code/3a_bimodality_coefficients.R"))
source(here("Code/3a_overlapping.R"))

```

-   Ontario Provincial Election Study, produced by [LISPOP](http://lispop.ca)
-   Available for public use at <https://doi.org/10.5683/SP3/VFGL0E> <!-- DOI doesn't work-->
-   Online consumer sample from SSI, weighted by gender, age and education
-   Campaign survey fielded between `r paste(format(min(on18$datestamp),"%B %d"),format(max(on18$datestamp),"%B %d, %Y"), sep="-")`
-   Rich set of questions on news consumption and social media usage
    -   observational, cross-sectional and self-reports, so ....


## Sample Validation

```{r}
#| label: tbl-vote-choice
#| tbl-cap: Vote intention of all respondents compared with actual vote results
data.frame(Party=c("Liberal Party of Ontario", "Progressive Conservative Party of Ontario", 
"New Democratic Party of Ontario", 
"Green Party of Ontario"), Vote=c(0.196, 0.405, 0.336, 0.046), Sample=rep("Election", 4))->vote_shares

on18 %>% 
  as_factor() %>%
filter(!is.na(partyvote2018)) %>% 
group_by(partyvote2018) %>% 
summarize(n=n()) %>% 
mutate(Vote=(n/sum(n)), Sample=rep("OPES", nrow(.))) %>% 
select(-n, Party=partyvote2018) %>% 
bind_rows(., vote_shares) %>% 
mutate(Vote=scales::percent(Vote, accuracy=1)) %>% 
pivot_wider(., names_from="Sample", values_from="Vote") %>% 
kable()
```
## Headline Findings

-   Partial evidence for polarization by social media and online news consumption.
-   heavy social media use \rightarrow slight increase in affective polarization
-   online news media consumption \rightarrow slight increase in policy polarization
-   Contra Dubois and Blank [-@duboisEchoChamberOverstated2018] political interest is directly, positively correlated with affective and policy polarization. 
  - Social media use and digital news consumption moderates this relationship; Polarization by social media use and digital news consumption mostly occurs amongst low interest voters


## Key Independent Variables

-   **Primary media source**: respondents were asked how they primarily received news about the 2018 Ontario election
    -   Respondents grouped into **legacy** (TV, radio, print newspapers) **online** (news aggregators, blogs, online newspaper) and **social media** (facebook or twitter feeds); otherwise **mixed**.
-   **Social media usage**: How often respondents use social media from **never** to **multiple times a day**
    -   No reference to news consumption


## Key Dependent Variables

**Affective polarization** 

- Increasing dislike of out-partisans 
   - Problem with multi-party democracies; weight by vote share [@wagnerAffectivePolarizationMultiparty2021]. 
   - **Calculates the average weighted difference between a respondent's average like score and their like score for each party, weighted by party vote share**


## Key Dependent Variables

**Policy polarization**

-   Index of 11 policy items (see Appendix in the paper).
-   Two measures of policy polarization (See [@lelkesMassPolarizationManifestations2016; @levenduskyRedStatesVs2011] 
   
   - bimodality coefficient to examine bimodality of a distribution of policy preferences [@pfisterGoodThingsPeak2013]. 
  
  - the overlap coefficient to examine overlap between two distributions of policy preferences [@pastoreMeasuringDistributionSimilarities2019].



<!-- <!--:::: {.columns} -->

<!-- ::: {.column width="50%"} -->

<!-- **Bimodality Coefficient** -->

<!-- $$ -->

<!-- BC = \frac{s^2 + 1}{k + 3 *\frac{(n-1)^2}{(n-2)(n-3)}} -->

<!-- $$ -->

<!-- ::: -->

<!-- ::: {.column width="50%"} -->

<!-- **Overlap Coefficent** -->

<!-- $$ -->

<!-- OVL = \int^{+\infty}_{-\infty}|f(x) - g(x)|dx -->

<!-- $$ -->

<!-- ::: -->

<!-- ::::-->

## Affective Polarization and Primary Media Source

```{r}
#| fig-height: 8
WAP_models %>%
  filter(Model_Name %in% c("Model 1", "Model 2", "Model 3", "Model 14")) %>%
  filter(Variable=="Primary_media") %>%
  #Filter out the interaction model
  #filter(Model_Name!="Model 8") %>%
  filter(str_detect(term, "Intercept", negate=T))  %>%
  ggplot(., aes(x=estimate, y=fct_relevel(term, "Social_Media", "Online", "Mixed", after=10)))+facet_wrap(~~fct_relevel(Model_Name, "Model 14",  after=3))+
  geom_pointrange(aes(xmin=estimate-(1.96*std.error), xmax=estimate+(1.96*std.error)))+geom_vline(xintercept=0, linetype=2, col="red")+labs(y="Variable", x="Coefficient") + theme_bw(base_size = 24)
```

## Affective Polarization and Social Media Usage

```{r}
#| fig-height: 8
WAP_models %>%
    filter(Model_Name %in% c("Model 1", "Model 2", "Model 3", "Model 14")) %>%
  filter(Variable=="Social_Use2") %>%
    #filter(Model_Name!="Model 12") %>%
  filter(str_detect(term, "Intercept", negate=T)) %>%
  ggplot(., aes(x=estimate, y=fct_relevel(term, "Never", "Several times a day","About once a day","Several times a week","About once a week", "Less than once a week", after=10)))+facet_wrap(~fct_relevel(Model_Name, "Model 14",  after=3))+
  geom_pointrange(aes(xmin=estimate-(1.96*std.error), xmax=estimate+(1.96*std.error)))+geom_vline(xintercept=0, linetype=2, col="red")+labs(y="Variable", x="Coefficient") + theme_bw(base_size=24)
```

## Interaction of social media use, interest and affective polarization

```{r}
#| label: fig-interactions
#| fig-cap: Predicted values of affective polarization by selected levels of social media use and interest in politics. 
#| fig-width: 5
#| fig-height: 3


plot_predictions(model6_su, 
                 condition=list("Interest", Social_Use2=c("Never", "About once a day", "Several times a day")), vcov=F)+
aes(linetype=Social_Use2, col=Social_Use2)+labs(y="Predicted WAP", col="Social Media Use")+theme_bw()+theme(legend.position="bottom")+
guides(col=guide_legend(ncol=2))


```



## Policy Polarization and Primary Media Source - Bimodality Coefficient

```{r}
library(gt)
on18 %>%
  group_by(Primary_media) %>%
  summarize(Bimodality=bimodality_coefficient(policy_polarization)) %>%
   mutate(Bimodality=round(Bimodality,2)) %>%
  gt()
```



## Policy Polarization and Primary Media Source - Bimodality Coefficient

```{r}

on18 %>%
  group_by(Interest_cat,Primary_media, ) %>%
  filter(!is.na(Interest_cat)) %>%
  summarize(Bimodality=bimodality_coefficient(policy_polarization)) %>%
rename(Interest=Interest_cat) %>%
  mutate(Bimodality=round(Bimodality,2)) %>%
gt()
```


## Policy Polarization and Social Media Usage - Bimodality Coefficient

```{r}

on18 %>%
  group_by(Social_Use2) %>%
  summarize(Bimodality=bimodality_coefficient(policy_polarization)) %>%
  filter(!is.na(Social_Use2)) %>%
#arrange(., desc(Bimodality)) %>%
  gt() %>%
fmt_number(., columns=2, decimals=2)
```

## Policy Polarization - Overlap Coefficent

```{r}
#theme_set(theme_bw(base_size=48))
library(cowplot)
ggpubr::ggarrange(overlap_legacy+theme_bw(base_size=24), overlap_online+theme_bw(base_size=24), overlap_smedia+theme_bw(base_size=24), overlap_mixed+theme_bw(base_size=24))
```


## Policy Polarization - Overlap Coefficent

```{r}
ggpubr::ggarrange(overlap_often+theme_bw(base_size=24), overlap_rarely+theme_bw(base_size=24), ncol = 1)
```

## Discussion & Conclusion

-   Some mild direct evidence for polarization
-   Interacts with interest in politics even after controlling for partisanship.
-   Consistent with Merkely (2022) who finds moderate increases in affective polarization, albeit we find increases amongst those with low levels of political interest
-  Inconsistent with Merkely (yesterday) who finds limited role for digital news consumption increasing affective polarization
-   Contra Dubois and Blank (2018), political interest \rightarrow polarization, social media usage and digital media consumption moderates this relationship.

## References {.allowframebreaks}

\footnotesize
