---
title: "Social Media Use And Polarization In Ontario: 2018 Case Study"
format: pdf
author: 
  - Rafael Campos-Gottardo
  - Simon Kiss
editor: visual
bibliography: Polarization.bib
---

## Introduction


```{r}
#| label: setup
#| echo: false
library(knitr)
opts_knit$set(root.dir=rprojroot::find_rstudio_root_file())
#opts_chunk$set(warning=F,message=F)
```

```{r}
#| label: import
#| echo: false
#| warning: false
#| message: false
#| results: 'hide'
#| cache: true
source("Code/3_polarization.R")
```
Will this work? We hope so.

```{r}
#| label: primary-media-interest-policy
primary_media_interest_policy
```

## Social Media And Polarization

- Define polarization, affective and policy
- Summary state of knowledge from big papers
  - Against [@boxellInternetCausingPolitical2017] and [@bakshyExposureIdeologicallyDiverse2015]
  - Try to review [@tuckerSocialMediaPolitical2018, @bavelHowSocialMedia2021,@kubinRoleSocialMedia2021]
- Summary of existing knowledge in Canada
  - Summarize [@duboisEchoChamberOverstated2018]
  - Others? 

## Ontario 2018 Summary

## Summary of OPES, Methods

- Sampling strategy
- Questions used to construct variables

### Measuring Polarization

- Affective polarization
- Policy polarization

## Results

## Discussion

## Conclusion

---

# References
