---
title: "Social Media Use and Polarization In Ontario: 2018 Case Study"
format: 
  pdf:
    number-sections: true
    colorlinks: true
    keeptex: true
    include-in-header: 
      text: |
          \usepackage{booktabs}
          \usepackage{siunitx}
          \usepackage{setspace}\doublespacing
          \newcolumntype{d}{S[
              input-open-uncertainty=,
              input-close-uncertainty=,
              parse-numbers = false,
              table-align-text-pre=false,
              table-align-text-post=false
           ]}
author: 
  - Rafael Campos-Gottardo
  - Simon Kiss
bibliography: Polarization.bib
link-citations: true
csl: apa.csl
---

# Introduction

```{r}
#| label: setup
#| echo: false
library(knitr)
opts_knit$set(root.dir=rprojroot::find_rstudio_root_file())
#opts_chunk$set(warning=F,message=F)
```

```{r}
#| label: import
#| echo: false
#| warning: false
#| message: false
#| results: 'hide'
#| cache: true
source("Code/3_polarization.R")
```



```{r}
#| label: primary-media-interest-policy
#| echo: false
#primary_media_interest_policy
```

# Social Media And Polarization

## What is Polarization

|     In the literature on polarization there has been much debate on the relationship between ideology and partisanship. Scholars studying polarization, particularly in the United States, have contrasting views on the nature of modern political polarization. Some scholars argue that ideological or policy polarization has been increasing for both political elites and the electorate more broadly. Another group argues that partisan sorting has led parties to become more ideologically homogeneous, while party elites have become more polarized, which leads to the perception that the electorate is becoming more polarized. Finally, a third group argues that despite opposing partisans being ideologically similar they increasingly dislike and distrust members of the opposing party [@iyengarOriginsConsequencesAffective2019]. These divergences in understandings of political polarization have led studies on polarization to differentiate between what is called called ideological/policy polarization [@masonIdeologuesIssuesPolarizing2018] and affective polarization [@masonUncivilAgreementHow2018] in their analyses of the causes and consequences of political polarization. 

### Ideological Polarization
|       Ideological polarization typically refers to one of two main claims. First, that supports of opposing parties or camps [@bantelCampsNotJust2023] have distinct political beliefs, which is also referred to as policy polarization. Second, that partisans are increasingly supportive of left or right-wing ideology and place themselves on the left or right of the political spectrum. These competing definitions of ideological polarization have led to different conclusions on the scale of ideological polarization in the United States and other advanced industrialized democracies. 
|       Policy based polarization is used to describe the growing differences in the beliefs systems of the electorate since the 1970s in the United States and the 1990s in other Western democracies [@daltonModelingIdeologicalPolarization2021]. Studies have provided evidence that Americans are increasingly divided on various emotionally charged issues including abortion, gay marriage and religion.
|       The idea that the public is ideologically polarized has long been questioned in the literature on public opinion and voting behaviour. @converseNatureBeliefSystems1964 found that Americans did not engage 

### Affective Polarization

## The relationship between social media usage and polarization

|       It is often considered a forgone conclusion that social media usage is linked to increased levels of affective and policy polarization. Proponents of this idea argue that social media platforms create "echo chambers" or "filter bubbles" where users will only encounter ideas and policies that they already agree with @terrenEchoChambersSocial2021. Since the internet is a high choice media environment individuals can either choose to be exposed to diverse information or select media that reinforce their pre-existing opinions [@duboisEchoChamberOverstated2018]. On social media echo chambers occur because users shown ideologically agreeable materials as a result of these platforms' algorithms that show users content based on their past behaviour (Raynauld & Grennberg 2014). However, the empirical evidence supporting the existence of eco-chambers on social media has been mixed. Most studies in the United States find that Republicans and Democrats have similar media diets, both online and offline. Additionally, studies have found that there is some evidence that echo chambers exist on Twitter (now known as X) but not on Facebook. Notably, Terren et al. @terrenEchoChambersSocial2021 found that studies that used digital trace evidence found more evidence of echo-chambers and polarization than studies that used self-reported data. For example, using Facebook trace evidence @bakshyExposureIdeologicallyDiverse2015 find that most users have Facebook friends who belong to the other party and that individuals are exposed to cross cutting content. They also find that individual choices, not algorithms, are what determine if an individual engages with opposing viewpoints. Using self reported social media usage, @duboisEchoChamberOverstated2018 find that the majority of social media users, encounter information they disagree with, use multiple sources, and often attempt to confirm the information they are presented.

|      The evidence linking the presences of echo chambers to increased polarization is also unclear [@kubinRoleSocialMedia2021]. Some studies find that increased social media usage predicts increased polarization [@choInfluencingMyselfSelfReinforcement2018; @tuckerLiberationTurmoilSocial2017]. This relationship is also re-enforced by posting more on social media and sharing politically relevant content. However, other studies have found that social media usage has a small or no effect on polarization. Through a review of recent studies on the relationship between social media usage and polarization, @bavelHowSocialMedia2021 posit that although social media usage is unlikely to be the main driver of polarization but is often a key facilitator. A recent experimental evidence has found that individuals who de-activated their Facebook accounts became less polarized due to less exposure to political news and opinions. Other studies have demonstrated that the "echo chamber" aspect of social media is not what drives polarization. Instead, exposure to hyper-partisan messages from the opposing party leads to a significant increase in polarization among Republicans but not democrats. @boxellInternetCausingPolitical2017 observe that the individuals who are the most polarized are those who are the least likely to use social media. Specifically, they found that adults over 75 are becoming polarized at a faster rate than those under 40, while also being far less likely to use social media. Therefore, any account that links social media usage to increased polarization must also account for why individuals who do not use social media are becoming more polarized than social media users.

|       The evidence that social media usage leads to increased polarization in Canada is even more limited. There is evidence that affective polarization has been increasing in Canada [@johnstonAffectivePolarizationCanadian2019]. Additionally, there is are mixed conclusions with regards to policy polarization in Canada. One study that measured polarization using preferences towards redistributive policies found that Canada experienced a surge in partisan sorting between 1992 and 2015 [@kevinsGrowingApartPartisan2018]. However, another study that measured polarization using distribution based measures only found evidence of ideological polarization increasing after the 2019 Canadian election [@merkleyPolarizationEhIdeological2022]. Furthermore, none of these studies investigate the factors that have contributed to the rise of both affective and ideological polarization in Canada.

## Polarization and Online News Consumption

|      In contrast to the contested relationship between polarization and social media usage their is clearer evidence that increased online news consumption is linked to increased polarization. Early studies of online news media found that partisan online sources engage in greater partisan filtering of content than traditional news wires [@baumNewMediaPolarization2008]. Additionally, studies have found that individuals who consume news online are often more polarized than those who only use offline "legacy" news sources [@fletcherHowPolarizedAre2020]. @garimellaPoliticalPolarizationOnline2021 find that both the structure of news sites and the behaviour of online news consumers contributes to users only consuming news that reinforces their ideological priors. Specifically, they find that the online news environment allows users actively seek out news sources that are ideologically aligned with their prior beliefs.

# Ontario 2018 Summary

There are several reasons to think that the Ontario 2018 general election might feature some degrees of polarization. First, it was the fifth election since the Ontario Liberals were first elected in 2003. As with any government with such a long tenure, the government had endured the consequences of many decisions gone wrong. The provincial government's attempts to increase the role of renewable energy in the provincial electricity supply had lead to an increase in electricity rates. In 200X, the government suddenly cancelled the construction of two natural gas electricity plants in electorally-sensitive districts, incurring hundreds of millions of dollars in penalties. in 200X, it emerged that the provincial air transportation agency for patients had become embroiled in complicated contractual arrangements, incurring millions of dollars in liabilities. In 20015, a new provincial sex education curriculum was adopted, including some controversial material incurring the wrath of social conservatives. 

Moreover, the two primary party leaders in the election campaign, Kathleen Wynne for the Liberals and Doug Ford for the PRogressive Conservatives were individuals with traits and backgrounds that might incur distinct enmity from out-partisans. Wynne was female and lesbian, contributing to some of the hostility by social conservatives. Even more notable, Ford 

# Data and Methods

## The 2018 Ontario Post Election Study

The Laurier Institute for the Study of Public Opinion and Policy commissioned an online consumer sample of voters from SSI with quotas established to match the sample to the population on age, education and gender. It was fielded between the dates of `r `

## Indenpedent Variables

## Polarization Varaibles

### Affective polarization

We observe affective polarization using two commonly used measured to measure affective polarization in multiple party systems. The first first measure, developed by @wagnerAffectivePolarizationMultiparty2021, creates an individual affective polarization score for each respondent based on the *spread* party like scores for each of Ontario's four major political parties. Like scores for the parties are drawn from feelings thermometers that ask respondents to rate the parties on a scale from 0 to 5 (See the Supplemental Materials for full question wordings).\
For formally the Weighted Affective Polarization (WAP) equation measured spread for parties j and voters i: 
$$
    Spread_i = \sqrt{\sum^{P}_{p=1}v_p(like_{ip} - \overline{like_i})^2}
$$ 
where $v_p$ is the vote share of each party measured as proportion with a range of 0 to 1, and the mean affect scores weighted by party using the following equation:

$$
    \overline{like_i} = \sum^P_{p=1}(v_p * like_{ip})
$$ 

The WAP scores are then used in OLS regression with the independent variables discussed above to determine understand the relationship between social media usage, political interest and affective polarization.

### Policy polarization

In order to measure policy polarization we create ideological distributions based on 12 policy issues. All the measures were re-coded so that 0 indicates the most left-wing position and 1 indicates the most right-wing position. The items were then added to create an ideological distribution that includes all 11 items. In this 0 represents being the most left-wing on all issues and 11 represents being the most right-wing on all issues.

#### Bimodality Coefficent

First we use these distributions to calculate the bimodality coefficient (BC). The BC is calculated using the `bimodality_coefficient` function from the `mousetrap` package in `R` which employs the following formula:

$$
BC = \frac{s^2 + 1}{k + 3 * \frac{(n-1)^2}{(n-2)(n-3)}}
$$ where $s$ represents the skewedness of the distribution, $k$ represents the distributions excess kurtosis, and $n$ refers to the sample size. A $BC$ of 0 indicates a unimodal distribution and a $BC$ of 1 indicates a distribution is considered **completely** bimodal. A distribution is typical considered bimodal when $BC \geq 0.55$. In order to measure the relationship between social media usage and ideological polarization using the BC we calculate the BC for the ideological distributions of individuals who primarily receive news from legacy media, online media sources, social media, and those who primarily use a mix of the three. Additionally, we compare the BC for those who use social media regularly and those who do not.

#### Distinctiveness Coefficent

We also use the distinctiveness coefficient first used to measure levels of polarization @lelkesMassPolarizationManifestations2016 and @levenduskyRedStatesVs2011 which is typically referred to as the overlap coefficient (OVL). This measure compares the overlap of the ideological distributions of two groups using the following formula:

$$
OVL = \int^{+\infty}_{-\infty}|f(x) - g(x)|dx
$$ 
where $f(x)$ is the probability density function (PDF) of one ideological distribution and $g(x)$ is the PDF of another. We calculate the the OVL using the `overlap` function from the `bayestestR` package in `R` ##fix. We compare the distributions of those who voted for the Liberal and NDP parties (left-leaning parties) to those who voted for the Conservative Party (right-leaning party) for those primarily receive news from legacy media, online media sources, social media, and those who primarily use a mix of the three and for those who use social media regularly and those who do not. In order to compare

## Results

## Discussion

## Conclusion

\newpage

# References
